---
title: "5222project"
output:
  pdf_document: default
  html_document: default
date: "2025-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(nlme)
library(knitr)
library(broom.mixed)
library(dplyr)
```

```{r}
data <- read.table("alcohol.txt", header = TRUE) %>%
  na.omit()

hist(data$alcuse, 
     breaks = 20,           
     main = "Histogram of Alcohol Use (alcuse)", 
     xlab = "Alcohol Use (alcuse)")

plot(density(data$alcuse), 
     main = "Density Plot of Alcohol Use (alcuse)",  
     xlab = "Alcohol Use (alcuse)")    

```
Not sure if we need to categorizes the variable peer(like the 6 point scale said in the instruction or other kind of levels).
```{r}
data$peer <- round(data$peer)

data %>%
ggplot(aes(x = age, y = alcuse, color = as.factor(coa))) +
geom_point() +
stat_summary(aes(group = coa), fun = mean, geom="line") +
labs(title ="Mean trajectories by Coa",y = "Alcohaol use", x = "Age", color = "Alcoholic parents(Yes/No)")+
  theme_minimal()

data %>%
ggplot(aes(x = age, y = alcuse, color = as.factor(peer))) +
geom_point() +
stat_summary(aes(group = peer), fun = mean, geom="line") +
labs(title ="Mean trajectories by peer",y = "Alcohaol use", x = "Age", color = "Peer level")+
  theme_minimal()

data %>%
ggplot(aes(x = as.factor(age), y = alcuse, color = as.factor(coa))) +
geom_point() + 
geom_line(aes(group = id)) + 
labs(title ="Individual trajectories by Coa",y = "Alcohol use", x = "Age", color = "Alcoholic parents(Yes/No)")+
  theme_minimal()

data %>%
ggplot(aes(x = as.factor(age), y = alcuse, color = as.factor(peer))) +
geom_point() + 
geom_line(aes(group = id)) + 
labs(title ="Individual trajectories by peer",y = "Alcohol use", x = "Age", color = "Peer level")+
  theme_minimal()


data <- data %>%
  mutate(logalcuse = log(alcuse))

data %>%
ggplot(aes(x = as.factor(age), y = logalcuse, color = as.factor(coa))) +
geom_point() + 
geom_line(aes(group = id)) + 
labs(title ="Individual trajectories by Coa",y = "log Alcohol use", x = "Age", color = "Alcoholic parents(Yes/No)")+
  theme_minimal()

data %>%
ggplot(aes(x = as.factor(age), y = logalcuse, color = as.factor(peer))) +
geom_point() + 
geom_line(aes(group = id)) + 
labs(title ="Individual trajectories by peer",y = "log Alcohol use", x = "Age", color = "Peer level")+
  theme_minimal()


```
It seems quadratic pattern for the individual level time trajectories. 

Trying to fit quadratic trend and individual trajectories differed by parental alcoholism. 

data of alcuse contain 0s which are not positive so that the gamma distribution does not satisfy. Adjust with adding very small number.

```{r}
data <- data %>%
  mutate(agesq = age^2)

# data of alcuse contain 0s which are not positive so that the gamma distribution does not satisfy. Adjust with adding very small number.
data$alcuse <- data$alcuse + 0.00001


#linear trend (random intercept)
m_linear <- glmer(alcuse ~ male + peer + coa + age + age * coa + age * peer + (1|id), data = data, 
              family = Gamma(link = "log"))
summary(m_linear)



#quadratic trend (random intercept)
m_quadratic <- glmer(alcuse ~ male + peer + coa + age + agesq + age * coa + 
                 agesq * coa + age * peer + agesq * peer + (1|id), data = data, 
              family = Gamma(link = "log"))
summary(m_quadratic)

l1 <- logLik(m_linear)
l2 <- logLik(m_quadratic)
l1
l2


LRT <- (-2)*(l1-l2)
LRT
LRT <- as.numeric(LRT)
p_value <- 1-pchisq(LRT, df=3)
p_value
p_value <- as.numeric(p_value)

anova(m_linear, m_quadratic, test = "LRT") 

table1 <- tibble(model = c("linear trend", "quadratic trend"), 
                 loglikelihood = c(l1,l2), df = c(3," "), statistic = c(1.94, " "), p_value = c(0.58, " "))

#table 1
table1 %>%
  kable(caption = "likelihood ratio test of linear trend and quadratic trend",
        digits = 2)



# Extract fixed effects with tidy format
coef_table <- tidy(m_linear, effects = "fixed")

# Round and rename columns
coef_table <- coef_table %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  dplyr::select(- effect)

# table 2
coef_table %>%
  kable(caption = "linear trend output(fixed effect)")




# model peer as ordinal continuous
model_ctspeer <- glmer(alcuse ~ peer + coa + age + age * peer + (1|id), data = data, 
              family = Gamma(link = "log"))
summary(model_ctspeer)
l1 <- logLik(model_ctspeer)


data$peer <- as.factor(as.character(data$peer))

# model peer as level
model_levelpeer <- glmer(alcuse ~ peer + coa + age + age * peer + (1|id), data = data, 
              family = Gamma(link = "log"))
summary(model_levelpeer)
l2 <- logLik(model_levelpeer)
LRT <- (-2)*(l1-l2)
p_value <- 1-pchisq(LRT, df=4)
p_value

anova(model_ctspeer, model_levelpeer)

table3 <- tibble(model = c("Ordinal peer", "Categorical peer"), 
                 loglikelihood = c(l1,l2), df = c(4," "), statistic = c(1.25, " "), p_value = c(0.87, " "))

#table 3
table3 %>%
  kable(caption = "likelihood ratio test:using peer as ordinal variable vs categorical factor",
        digits = 2)



# Extract fixed effects with tidy format
coef_table <- tidy(model_ctspeer, effects = "fixed")

# Round and rename columns
coef_table <- coef_table %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  dplyr::select(- effect) %>%
  dplyr::select(- p.value) %>%
  dplyr::select(- statistic)

#table 4
coef_table %>%
  kable(caption = "final model output(fixed effect)")



```
We start from comparing two models, linear trend and quadratic trend over age(which include both age and age square term in the model). We use likelihood ratio test to test which one is preferred for our data. After the preferred one is chosen, we try to fit both the random intercept model and random slope model(include both random intercept and random slope). 

From the original dataset, predictor "peer" is a floating continuous variable. We round it to nearest integer(0 = "none", 5 = "all") in order to improve interpretability of group effect. Therefore, we can fit the model either by treating "peer" as categorical variable with 6 possible levels(we only have 3 levels(0, 1 and 2) in our data), or as the ordinal continuous variable without loss of interpretability. To compare which approach is better for our data, we applies likelihood ratio test. 

From the fixed effect, to see if the group*time effect exist(Is the time effect differed by group?), we test if the coefficients of (coa, age) interaction effect and (peer, age) interaction effect are statistical significantly different from zero. That means we conduct hypothesis tests with null hypothesis: beta5 equal to zero and beta6 equal to zero. 

After removing all non-significant terms, we refit the model and our final model is obtained. 

Note: we add a very small number(0.0001) on the response variable(alcuse). That's because some value of response variable are zero which violates the gamma assumption of positive value. Making such minor adjustment helps us to get a converged model. 




results:



```{r}

```

```{r}
cor(data$age, data$age^2)

data <- data %>%
  mutate(agec = age-mean(age))%>%
  mutate(agecsq = agec^2)

cor(data$agec, data$agec^2)



```




